A few stickleback suffice to transport adaptive alleles to new lakes

### 2018-12-12 Stickleback_Paper.tex
latest

The canonical model for the genetics of adaptation, first formulated by Fisher in the early part of the 20th century, involves the sequential fixation of new mutations. 
While it has proven valid in numerous studies in the field as well as in the lab, this model is now rightfully understood as incomplete for many species in nature that have more complicated population structures. 
A growing number of studies have identified instances of convergent or shared adaptive evolution in small demes that utilize standing genetic variation (SGV) found in the static population from which they split.
However, it is still widely unknown which variations of evolutionary processes such as gene flow, recombination, selection and mutation, promote the maintenance and re-use of such genetic polymorphisms.
How do complex geographic systems with a variety of selective pressures interact with these evolutionary forces to help or hinder the use of SGV in newly colonized subpopulations?

Recently, a large flood of data resulting from systems such as this come from population genomic studies in nature that have been powered by advances in sequencing technologies. 
One such organism is the threespine stickleback which has exhibited a long standing evolutionary history of migration-selection balance between marine and freshwater populations across the globe.
The rigidity of the species' ability to frequently prosper in newly created freshwater environments has made this fish a good model for understanding the genetic basis of adaptive evolution. 
The ancestral marine form of Stickleback has given rise to millions of independently derived populations in recently de-glaciated regions of the Northern Hemisphere.
Geographic isolation prevents direct migration between a large majority of these patches of freshwater populations,
yet we observe very similar phenotypes and many shared adaptive alleles often found to be identical by decent (IBD). 
Impressively, independent local adaptation of marine individuals to freshwater environments has been observed to take place in tens of generations, 
and the adaptive alleles are found in freshwater populations that have been geographically isolated since the end of the last ice age $\approx 13,000$ years ago (Cite Kristin).
For example, in 1964 the Great Alaskan Earthquake caused an uplift of Middleton island and in turn, introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of stickleback, 
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be parallel to freshwater stickleback that have been separated for over thousands of years. 
In these freshwater stickleback, the number of lateral plates are reduced and the opercle shapes shows
the same expansion of the dorsal region and reduction of the ventral region as observed in the large majority of freshwater demes.

But how can evolution occur at such a rapid pace? 
Waiting for new mutations to arise in each lake would take much longer, and genotypes being identical by state is even more improbable. 
This would seem to suggest that freshwater alleles are maintained in the marine individuals allowing the accelerated selection on SGV found in marine individuals which colonize the lake. 
The first clear example of the global reuse of SGV was the gene \textit{eda} which has been shown to be an important regulator for the number of lateral plates. 
While the low lateral plate version of this gene arose millions of years ago, it is found in freshwater ponds which have formed much more recently. 
More recently, population genomic studies employing genome-wide haplotype analyses has provided evidence that \textit{most} regions of the genome that distinguish marine-freshwater genetic differences share this pattern \citep{nelson2017ancient}.

@@

We find that only a few stickleback per generation, per lake, suffice to maintain freshwater alleles at an acceptable frequency in marine environments for efficient ``transportation" of the freshwater haplotype.
The selection on SGV at this level results in rapid local adaptation in $\approx 60$ generations, this follows closely to what we observe in nature (such as middleton island, cite this).
Surprisingly, we find that the the continued gene flow (subsequent migration) of marine individuals is less important than the initial amount of SGV in the initial generation marine individuals at time of colonization.
This suggests downstream gene flow from freshwater populations is key to this system of ``transportation".
We show that low levels of introgression also results in noise that often presents itself as $F_{st}$ peaks when scanning across the genome suggesting introgression plays a key role in inferring causal loci . 
With higher $M > 10$ individuals per lake per generation we see the impact of migration load, preventing the demes from fully adapting to their selective pressure.

@@

\begin{figure}
	\begin{center}
  		\includegraphics[width=0.6\linewidth]{GeographyFigure.pdf}
  		\caption{
            \textbf{Diagram of simulated populations:}
            a single, continuous, one-dimensional marine habitat (blue)
            is coupled to randomly mating ``lakes'' at discrete locations with dark green arrows representing migration patterns.
            %It is important to note there is no migration from any one lake to another, therefor, freshwater alleles must travel through the marine habitat as is often the case in nature
            After an initial period of 100K generations with 25 lakes,
            an additional 25 lakes are added (at the same set of locations) and populated with a copy of the \textit{marine} individuals
            to simulate the appearance of newly accessible freshwater habitats colonized by marine stickleback.
            The marine habitat, and each set of 25 lakes, each contain 5,000 individuals at all times.
			}
  		\label{fig:Geo}
	\end{center}
\end{figure}

@@

As shown in Figure~\ref{fig:MeanPhenotype}, local adaptation occurred at most migration rates with the exception of the highest migration at which half of each population was composed of migrants.
At lower migration rates (10 migrants per generation and below), populations adapted to local conditions.
As shown in Figures \ref{fig:phenotype_time} and \ref{fig:phenotype_time_supp},
freshwater and marine populations diverged until the trait means were close to the optimal values in each habitat. 
The establishment of new alleles in the lakes is visible in Figure \ref{fig:phenotype_time} as jumps in the mean trait value,
which move the trait by an amount of order 1 every few hundred generations.
Trait variation within each population was small compared to the difference between populations.
Across all parameter values, differences at around 16 commonly polymorphic sites 
(eight that shift the trait in each direction)
were responsible for most of the adaptive differences between freshwater and marine habitats.

As expected, increasing migration rate decreased differentiation between habitats. 
As seen in Figure~\ref{fig:Fst}, 
$F_{ST}$ between marine and freshwater habitats 
at neutral sites steadily declines as migration increases. 
However, local adaptation was still able to occur despite overall homogenization 
at up to 10 migrants per lake per generation.

@@

\paragraph*{Speed of adaptation}
Adaptation occurred much more quickly
at higher migration rates,
both in the old and new sets of lakes.
We measured this ``time to adaptation''
as the number of generations until 
average trait values in old and new lakes were within 0.5 of each other,
shown in Figure \ref{fig:TimeToAdaptation} for different rates of gene flow.
Adaptation of new lakes took over 18,000 generations at the lowest rate,
while at one migrant per lake per generation
new lakes managed to adapt in only 32 generations.

@@

At low migration rates, the \emph{initial} period of adaptation
takes roughly ten times longer for lakes as it does for the ocean.
This difference occurs because at low migration rates, 
adaptation occurs independently in each lake,
and the marine habitat has ten times as many individuals,
and therefore ten times the influx of new alleles than any one lake. 
At higher migration rates, greater mixing
allows lakes to share alleles instead of developing their own genetic basis of adaptation. 
As a first indication of this, Figure \ref{fig:Fst} 
shows that $F_{ST}$ between the ``original'' and ``introduced'' sets of lakes 
at effect mutations decreased with migration rate.

To investigate in more depth how locally adaptive alleles found in the original lakes
are shared between lakes, and how they spread to the new lakes,
we counted ``pre-existing freshwater adapted alleles''.
These we define for a particular generation 
to be any mutation with an effect on the trait
whose frequency is above 50\% in at least one original lake 
and below 50\% in the marine habitat.
Figure \ref{fig:MPFAI}A shows the distribution of the number of these alleles, 
across generations.
At the lowest migration rate (one migrant per 100 generations per lake), 
each lake has a private set of about 10 mutations
nearly fixed in that lake but not elsewhere:
new lakes acquire new adaptive alleles and so have none of these.
At one migrant per 10 generations per lake,
the original set of lakes each developed private alleles,
but these are present at low frequency in the ocean,
and so new lakes adapt 
mostly using a subset of this large repertoire of standing variation.
As migration rate increases past this, 
the number of pre-existing freshwater adapted alleles declines
because alleles move between populations by migration
before they can appear by new mutation,
while the frequency of these alleles in the ocean
stays relatively constant.

Figure~\ref{fig:MPFAI}B shows the distribution of 
the mean percentage of currently-defined freshwater adapted alleles 
that each genome in each of the populations carries, 
averaged across time and individuals. 
If all individuals across lakes carried the same set of alleles determining their trait value, 
this would be 100\%. 
At the lowest migration rates,
each genome in the original lakes have almost exactly~$1/25^{th}$ 
of the total number of pre-existing freshwater adapted alleles -- 
this is because each of the 25 lakes has adapted with a unique set of alleles. 
Since these are \emph{pre-existing} alleles, the value is zero for introduced lakes.
Figure \ref{fig:MPFAI}A shows us that at $0.1$ migrants per lake per generation and above,
the average individual across the new lakes has nearly the same amount of 
pre-existing freshwater adapted alleles as individuals across the old lakes.
As expected, the genetic basis of the freshwater phenotype 
seems to simplify as migration increases --
higher rates of migration allow
adaptive alleles of higher effect to travel more efficiently through the population,
even though they are deleterious in the ocean.

The numbers of Figure~\ref{fig:MPFAI} suggest that
the dramatic increase in speed of local adaptation we observed above occurs because
higher gene flow between populations allows sharing of freshwater alleles between populations.
We confirmed this by using recorded tree sequences to identify the origin of each trait-affecting
allele common in the new lakes at time of adaptation,
as defined in the Methods.
Figure~\ref{fig:Origin} shows that at the lowest rate of gene flow,
the majority of adaptive alleles are derived from de novo mutation.
As gene flow increases, a larger fraction of adaptive alleles 
derive from pre-existing variation in the marine population at the time of introduction.
In other words, 
greater mixing at higher migration rates allows lakes to share alleles
instead of developing their own genetic basis of adaptation.

@@

At first, increased migration allows sharing of adaptive alleles between lakes, 
but at the highest migration rate, 
the constant influx of alleles between the habitats creates substantial migration load. 
The rate at which migration load becomes substantial,
10 migrants per lake per generation,
only replaces 5\% of each population each generation with migrants from the other habitat, 
but this is sufficient to shift the mean trait values to nearly half their optimal values, 
as seen in Figure \ref{fig:MeanPhenotype}.

@@

Now we take a closer look at the genomic architecture of local adaptation between the two habitats. 
Do the alleles underlying trait differences cluster along the genome?
Do measures of local differentiation identify the causal loci? 
Figure \ref{fig:Fst} shows plots along the genome of per-locus $F_{ST}$ values between habitats.
(Note that we are pooling freshwater habitats; a single lake would provide substantially less power.)
\plr{need to know where the FAA are in these figures}
Higher migration rates showed more distinct $F_{ST}$ peaks over polymorphic loci 
underlying trait differences between the habitats. 
``Background'' levels of $F_{ST}$ increase as gene flow decreases,
swamping out this signal until the regions under selection are indistinguishable. 
This is likely due to two reasons: 
first, stronger genetic drift with less migration leads to higher background $F_{ST}$, and 
second, greater sharing of adaptive alleles providing a shared signal across populations.

>>>>

The canonical model for the genetics of adaptation, first formulated by Fisher in the early part of the 20th century, involves the sequential fixation of new mutations. 
While it has proven valid in numerous studies in the field as well as in the lab, this model is now rightfully understood as incomplete for many species in nature that have more complicated population structures. 
A growing number of studies have identified instances of convergent or shared adaptive evolution in small demes that utilize standing genetic variation (SGV) found in the static population from which they split.
However, it is still widely unknown which variations of evolutionary processes such as gene flow, recombination, selection and mutation, promote the maintenance and re-use of such genetic polymorphisms.
In addition, we would like to know how complex geographic systems, with a variety of selective pressures, interact with these evolutionary forces to help or hinder the use of SGV in newly colonized subpopulations.

Recently, a large flood of data resulting from evolutionary systems of organisms have come from population genomic studies powered by advances in sequencing technologies. 
One such organism is the threespine stickleback. 
Many studies of species have exhibited a long standing evolutionary history of migration-selection balance between marine and freshwater populations across the globe.
The rigidity of the stickleback's ability to frequently prosper in newly created freshwater environments has made this fish a good model for understanding the genetic basis of adaptive evolution. 
The ancestral marine form of stickleback has given rise to millions of independently derived populations in recently de-glaciated regions of the Northern Hemisphere.
While geographic isolation prevents direct migration between a large majority of these patches of freshwater populations,
They appear to exhibit very similar phenotypes, as well as many shared, adaptive alleles often found to be identical by decent (IBD). 
Impressively, independent local adaptation of marine individuals to freshwater environments has been observed to take place in tens of generations, 
and the adaptive alleles are found in freshwater populations that have been geographically isolated since the end of the last ice age $\approx 13,000$ years ago (Cite Kristin).
For example, in 1964 the Great Alaskan Earthquake caused an uplift of Middleton island and in turn, introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of stickleback, 
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be parallel to freshwater stickleback that have been separated for over thousands of years. 
In these freshwater stickleback, the number of lateral plates are reduced and the opercle shapes shows
the same expansion of the dorsal region and reduction of the ventral region as observed in the large majority of freshwater demes.

But how can evolution occur at such a rapid pace? 
Waiting for new mutations to arise in each lake would take much longer, and the genotypes being identical by state is even more improbable. 
This would seem to suggest that freshwater alleles are maintained in the marine individuals allowing the accelerated selection on SGV found in marine individuals which colonize the lake. 
The first clear example of the global reuse of SGV was the gene \textit{eda} which has been shown to be an important regulator for the number of lateral plates. 
While the low lateral plate version of this gene arose millions of years ago, it is found in freshwater ponds which have formed much more recently. 
Contemporary population genomic studies employing genome-wide haplotype analyses has provided evidence that \textit{most} regions of the genome that distinguish marine-freshwater genetic differences share this pattern \citep{nelson2017ancient}.

@@

We find that only a few stickleback per generation, per lake, 
suffice to maintain freshwater alleles from a (global?) haplotype at an acceptable frequency in marine environments for efficient ``transportation" of the freshwater haplotype.
The selection on SGV at this level results in rapid local adaptation in $\approx 60$ generations, this follows closely to what we observe in nature (such as middleton island, cite this).
Surprisingly, we find that the the continued gene flow (subsequent migration) of marine individuals is less important than the initial amount of SGV in the initial generation marine individuals at time of colonization.
This suggests downstream gene flow from freshwater populations is key to this system of ``transportation".
We show that low levels of introgression also results in noise that often presents itself as $F_{st}$ peaks when scanning across the genome suggesting introgression plays a key role in inferring causal loci . 
With higher $M > 10$ individuals per lake per generation we see the impact of migration load, preventing the demes from fully adapting to their selective pressure.

@@

\begin{figure}
	\begin{center}
  		\includegraphics[width=0.6\linewidth]{GeographyFigure.pdf}
  		\caption{
            \textbf{Diagram of simulated populations:}
            a single, continuous, one-dimensional marine habitat (blue)
            is coupled to randomly mating ``lakes'' at discrete locations with dark green arrows representing migration patterns.
            After an initial period of 100K generations with 25 lakes,
            an additional 25 lakes are added (at the same set of locations) and populated with a copy of the \textit{marine} individuals
            to simulate the appearance of newly accessible freshwater habitats colonized by marine stickleback.
            The marine habitat, and each set of 25 lakes, each contain 5,000 individuals at all times.
			}
  		\label{fig:Geo}
	\end{center}
\end{figure}

@@

As shown in Figure~\ref{fig:MeanPhenotype}, local adaptation occurred at migration rates of $m \le 10$.
% with the exception of the highest gene flow at which half of each population was composed of migrants.
%At lower migration rates (10 migrants per generation and below), populations adapted to local conditions.
Figures \ref{fig:phenotype_time} and \ref{fig:phenotype_time_supp}
show freshwater and marine populations diverging until the trait means were close to the optimal values in each habitat. 
The establishment of new alleles in the lakes is visible in the first few thousand generations of Figure \ref{fig:phenotype_time} as jumps in the mean trait value.
%which move the trait by an amount of order 1 every few hundred generations.
%Trait variation within each population was small compared to the difference between populations.
At the lowest rate of gene flow, $m = 0.01$, differences at $\approx 20$ commonly polymorphic sites 
($\approx 10$ that shift the trait in each direction)
were responsible for most of the adaptive differences between freshwater and marine habitats.
In Figure \ref{fig:MPFAI}A, we can see that more gene flow simplified the genetic basis. %does this belong here?
As expected, increasing migration rate decreased differentiation between habitats. 
As seen in Figure~\ref{fig:Fst}, 
$F_{ST}$ between marine and freshwater habitats 
at neutral sites steadily declines as migration increases.

@@

\paragraph*{Speed of adaptation}
Adaptation occurred much more quickly
at higher migration rates,
both in the old and new sets of lakes.
We measured this ``time to adaptation''
as the number of generations until 
average trait values in old and new lakes were within 0.5 of each other,
shown in Figure \ref{fig:TimeToAdaptation} for different rates of gene flow.
Adaptation of new lakes took over 18,000 generations at the lowest rate of $m = 0.01$,
while at $m = 1$ migrant per, lake per generation,
new lakes managed to adapt in just under 60 generations. 
%This dramatic decrease in time to local adaptation is suggestive of just how quickly
%selection can act on standing variation. 
%It also suggests a rough threshold of gene flow required 
%to efficiently select upon the alleles present in the marine individuals.

@@

At low migration rates, the \emph{initial} period of adaptation
takes roughly 25 times longer for lakes than it does for the ocean.
This difference occurs because at low migration rates,
all subpopulation must wait for novel mutations to arise in order to adapt.
Because the marine is one continuous population with 25 times more individuals than any one lake, 
there is a much larger influx of new mutations to be selected upon. 
At higher migration rates, greater mixing
allows the initial lakes to share alleles instead of developing their own genetic basis of adaptation. 
As a first indication of this, Figure \ref{fig:Fst} 
shows that $F_{ST}$ between the ``original'' and ``introduced'' sets of lakes 
at effect mutations decreased with migration rate.
%We are interested in this shared genetic basis of adaptation because it is commonly thought that this is
%a requirement for the efficient ``transportation" of the freshwater haplotype in to new lakes \citet{schluter2009genetics}

To investigate in more depth how locally adaptive alleles found in the original lakes
are shared between lakes, as well as how they spread to the new lakes,
we defined and tracked the distribution of ``pre-existing freshwater adapted alleles'' at the beginning of each generation.
To be considered in this category, a mutation must participate in the genetic basis of any freshwater adaptation.
Concretely, these are any non-neutral mutations whose frequency is above 50\% in at least one original lakes and below 50\% in the marine habitat.
Figure \ref{fig:MPFAI}A shows the distribution of the number of these alleles across generations.
At $m = 0.01$, each lake has a private set of about 10 mutations nearly fixed in that lake but not elsewhere:
the new lakes acquire new adaptive alleles and therefor have little to none of these adaptive alleles.
At $m = 0.1$, we again observe the original lakes adapting nearly independently from each other.
However, in Figure \ref{fig:MPFAI}A we can see the average marine individual is shown to hold $\approx 2$ pre-existing freshwater adapted alleles. 
This being the case, we can see new lakes adapt using a subset of this large repertoire of standing variation deriving from a multitude of independently derived haplotypes.
As migration rate increases past this, %we can see migration selection balance keeps the average capacity for marine individuals to carry freshwater alleles at about 2.
we can see in Figure \ref{fig:MPFAI}C that the total number of pre-existing freshwater adapted alleles declines.
%while the number of alleles which make up the genetic basis of adaptation in freshwater individuals remains relatively similar.
%Although the genetic basis seems to simplify slightly due to more , this 
This suggests that the alleles move between populations by migration 
before selection acts on new mutation for neighboring lakes.
Interestingly, the frequency of these alleles in the ocean
stays relatively constant across the reasonable rates of gene flow.

Figure~\ref{fig:MPFAI}B shows the distribution, through time, of 
the mean percentage of currently-defined freshwater adapted alleles 
that each genome in each of the populations carries.
If all individuals across lakes carried the same set of alleles determining their trait value, 
this would be 100\%. 
At the lowest migration rate $m = 0.01$,
each genome in the original lakes have almost exactly~$1/25^{th}$ 
of the total number of pre-existing freshwater adapted alleles -- 
this is because each one of the 25 lakes has adapted with a unique set of alleles. 
Since these are \emph{pre-existing} alleles, the value is zero for introduced lakes.
Figure \ref{fig:MPFAI}A shows us that at $0.1$ migrants per lake per generation and above,
the average individual across the new lakes has nearly the same amount of 
pre-existing freshwater adapted alleles as individuals across the old lakes.
As expected, the genetic basis of the freshwater phenotype 
seems to simplify as migration increases --
higher rates of migration allow
adaptive alleles of higher effect to travel more efficiently through the population,
even though they are deleterious in the ocean.

The numbers of Figure~\ref{fig:MPFAI} suggest that
the dramatic increase in speed of local adaptation we observed above occurs because
higher gene flow between populations allows sharing of freshwater alleles between populations.
%however, it is not enough for the marine population to hold bits and peices from independently formed haplotypes,
%For selection to act on SGV on the scale we observe in nature, the gene flow 
%initial generation to be able to rebuild 
We confirmed this by using recorded tree sequences to identify the origin of each trait-affecting
allele, in each individual, common in any of the new lakes at time of adaptation
as defined in the Methods.
Figure~\ref{fig:Origin} shows that at the lowest rate of gene flow the majority of adaptive alleles are derived from \emph{``de novo"} mutation.
As gene flow increases, a larger fraction of adaptive alleles 
derive from pre-existing variation in the marine population at the time of introduction.
In other words, 
greater mixing at higher migration rates allows lakes to share alleles
instead of developing their own genetic basis of adaptation.

@@

While increased migration allows sharing of adaptive alleles between lakes, 
at $m = 10.0$ the constant influx of alleles between the habitats creates substantial migration load. 
The rate at which migration load becomes substantial,
10 migrants per lake per generation,
only replaces 5\% of each population each generation with migrants from the other habitat, 
but this is sufficient to shift the mean trait values to nearly half their optimal values, 
as seen in Figure \ref{fig:MeanPhenotype}.

@@

Now we take a closer look at the genomic architecture of local adaptation between the two habitats. 
%Do the alleles underlying trait differences cluster along the genome?
Do measures of local differentiation identify the causal loci? 
Figure \ref{fig:Fst} shows plots along the genome of per-locus $F_{ST}$ values between habitats.
(Note that we are pooling freshwater habitats; a single lake would provide substantially less power.)
%\plr{need to know where the FAA are in these figures}
Higher migration rates showed more distinct $F_{ST}$ peaks over polymorphic loci 
underlying trait differences between the habitats. 
``Background'' levels of $F_{ST}$ increase as gene flow decreases,
swamping out this signal until the regions under selection are indistinguishable. 
This is likely due to two reasons: 
first, stronger genetic drift with less migration leads to higher background $F_{ST}$, and 
second, greater sharing of adaptive alleles providing a shared signal across populations.
